Synthesis of Rules for Network Intrusion Detectors

### 2019-10-23 paper/paper.tex
pass on examples and technique
pass

Suricata supports a total of \totoptions\ options covering
\numproto\ distinct protocols. Of these total of options,
\totoptionsrelevant\ are relevant to capture traffic (see
Figure~\ref{fig:distribution-rules-protocol}). Table~\ref{table:rules}
shows a sample of the relevant options of Suricata. Column ``\#''
shows the option number, column ``Name'' shows the name of the
option, column ``\#~Params'' shows the number of additional parameters
associated with a given option, and column ``Description'' presents a
short description of the option purpose.

@@

\begin{figure}[t]
  \lstinputlisting[language=C,numbers=none,keywords={flags,threshold}]{synflood.suricata}
  \caption{Suricata rule for SYN Flood Attacks.}
  \label{fig:synflood-example}
\end{figure}

@@

The SYN Flood attack is a denial-of-service attack that exploits a vulnerability in the TCP/IP handshake
to establish a TCP connection~\cite{cloudfare-synflood}. The handshake
works as follows in normal circumstances. First, a client sends a SYN
packet to the server, requesting a connection. Second, the server
responds with a SYN-ACK packet to the client. Third, the client
responds with an ACK message and the connection is established. Aware
of the protocol, an attacker sends multiple SYN packets to different
ports of a server, often using fake IP addresses. Then, after the
server responds with a SYN-ACK packet, the client keeps sending other
SYN packets to avoid the connection to time out. Without proper
protection, the server accepts these malicious requests and eventually
legitimate requests cannot be satisfied due to resource exhaustion.
Figure~\ref{fig:synflood-example} shows a Suricata rule that captures
this attack. This rule was obtained from the \Fix{...what source...}
The relevant options in the rule appear in bold---the option
\CodeIn{flags: S,12}, identifies a SYN packet in a TCP packet and the
option \CodeIn{threshold: type both, track by\_dst, count 5000,
  seconds 5} indicates that a high volume of such packets should be
requested in a short period of time.

@@

\Fix{---------------------------------------------------
  estou aqui}

@@

\vspace{1ex}
\noindent\textbf{Overview.}~\tname{} synthesizes rules in two
steps. First, it produces a potentially overspecified rule that
captures the malicious traffic provided on input. In this step,
\tname\ \emph{maximizes} the number of options in the rule that
captures the positive traffic. It extracts tokens from the malicious
message and use them to define the initial rule. Furthermore, it
searches for additional rule option (\eg, \Fix{xxx}) that are not
found in the message data but are satisfied by the traffic. We say
that this rule overfits the problem as it clearly satisfy the success
criterion of only capturing the positive traffic, however it is unable
to generalize to other attacks of the same kind with different
data. In the second step, \tname{} uses negative examples to
\emph{minimize} the rule, discarding unnecessary options. Note that
any subset of options in the rule obtained in this stage captures the
positive traffic by construction. \Fix{...elaborate...}

\subsection{Step 1: Rule Maximization}
\pgfplotsset{width=6cm,compat=1.8}
\pgfplotsset{every tick label/.append style={font=\tiny}}

The goal of the first step is to synthesize a rule with the maximum
number of options that satisfy the malicious traffic. A rule is
maximal if no other option can be added without missing the malicious
traffic.

\begin{wrapfigure}[16]{r}{0.5\textwidth}
  \centering
  \vspace{-5ex}
  \scalebox{1.05}{
    \begin{tikzpicture}
      \begin{axis}[
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
%          ylabel={\#rules},
          symbolic x coords={gen, http, icmp, udp, tcp},
          xtick=data,
          nodes near coords,
          every node near coord/.append style={font=\scriptsize},        
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(gen,70) (http,34) (icmp,94) (udp,84) (tcp,43)};
        \addplot coordinates {(gen,50) (http,20) (icmp,134) (udp,90) (tcp,123)};    
        \legend{relevant, irrelevant}
      \end{axis}
    \end{tikzpicture}
  }
  \caption{\label{fig:distribution-rules-protocol}Distribution of
    number of rules per protocol and relevance for triggering action.}
\end{wrapfigure}
It is important to note that options can be specific to a given
protocol or general and they can influence or not triggering an
action. The purpose of rule options \CodeIn{msg}, for instance, is to
print on the output an informative message indicating what kind of
intrusion has been
observed. Figure~\ref{fig:distribution-rules-protocol} shows the
distributions of these options per protocol and their relevance for
triggering rule actions. We obtained this information
by\Fix{...explicar como...} \Fix{eh um pouco suspeito ter tantas regras
  irrelevantes...}This plot indicates that the problem search
space is defined based on the protocol of the malicious traffic. For
instance, if \tname{} detects by analyzing the format of network packets that the attack involves \CodeIn{http}
traffic, then 34 different rule options will be considered in the
creation of the rule.

To infer options, \tname{} parses the protocol message, looking for
fields in the message associated with options. For example, fields
\Fix{...and...}  can be used to determine the option
\Fix{...}. Unfortunately, inferring options at the field granularity
level is insufficient to capture \CodeIn{content} options (see
Section~\ref{sec:example-suricata-rules}), which refer to parts of the
payload of the message and cannot be directly extracted directly from
the fields. Handling this case is critically important as most rules
contain \CodeIn{content} options. Figure~\ref{fig:distribution-contents} shows
the histogram of number of these options per rule (up to 10) for the
Suricata public rule database \Fix{cite}. Rules without \CodeIn{content}
options is an exception and half of the rules contain at least two
\CodeIn{content} options.
\begin{wrapfigure}[14]{r}{0.56\textwidth}
  \centering
  \vspace{-3ex}
  \scalebox{1.2}{  
    \begin{tikzpicture}
      \begin{axis}[
          bar width=6pt,
          scaled ticks=false,
          tick label style={/pgf/number format/fixed},
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          symbolic x coords={0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10},
          xtick=data,
          nodes near coords,
          every node near coord/.append style={font=\tiny},
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(0,25) (1,10222) (2,4956) (3,5589)
          (4,1651) (5,1457) (6, 729) (7, 471) (8, 313) (9, 137) (10, 126)};
      \end{axis}
    \end{tikzpicture}
  }
  \caption{\label{fig:distribution-contents}Histogram of number of
    \CodeIn{content} options per rule (up to 10).}
\end{wrapfigure}
To find the values associated with these options, \tname{} splits the
payload of the malicious message in tokens using natural language
delimiters, such as $\backslash$t, $\backslash$n, $\backslash$r, and
spaces. The rationale for this decision is that we observed, by
inspecting existing rules, that \CodeIn{content} options typically refer to
sequence of characters in the payload of messages separated by these
delimiters. \Fix{describe extraction of options not in the
  message. give examples! search only makes sense if this exists!}

@@

\Fix{...............estou aqui}

%% \begin{figure*}[t]
%% \centering
%% \includegraphics[trim=10 360 182 130,clip,width=0.95\textwidth]{figs/nids-workflow}
%% \caption{\tname\ workflow.}
%% \label{fig:overview}
%% \end{figure*}

Figure~\ref{fig:overview} shows the workflow of \tname{} as a pipeline
of two components, appearing numbered in the figure. Each component
implements one of the steps mentioned above. The goal of the first
component is to produce a rule that captures the negative traffic.
First, it initializes the rule with random values. Only options
associated with the used protocol are included in the rule
representation. Then, \tname\ optimizes the rule options until
\suri\ is able to capture the attack associated with the negative
traffic.  \tname\ unsuccessfully terminates at this point if it cannot
capture the attack. The second component takes as input the rule
produced by the first component and tries to minimize that rule. Note
that any subset of options would capture the negative traffic---as all
options in the rule need to be satisfied for the negative traffic to
be captured---but it can also capture positive traffic. This component
systematically discards options from the rule encoding until no
positive traffic is captured.

>>>>

Suricata supports a total of \totoptions\ options covering
\numproto\ distinct protocols. Of these \totoptions,
\totoptionsrelevant\ options are relevant to capture traffic
\begin{wrapfigure}[17]{r}{0.52\textwidth}
  \centering
  \vspace{-4ex}
  \scalebox{.75}{
    \begin{tikzpicture}
      \begin{axis}[
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
%          ylabel={\#rules},
          symbolic x coords={gen, http, icmp, udp, tcp},
          xtick=data,
          nodes near coords,
          every node near coord/.append style={font=\scriptsize},        
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(gen,70) (http,34) (icmp,94) (udp,84) (tcp,43)};
        \addplot coordinates {(gen,50) (http,20) (icmp,134) (udp,90) (tcp,123)};    
        \legend{relevant, irrelevant}
      \end{axis}
    \end{tikzpicture}
  }
  \caption{\label{fig:distribution-rules-protocol}Distribution of
    number of options per protocol and relevance for triggering action.}
\end{wrapfigure}
Options can be specific
to a given protocol or general and they can influence or not
triggering an action. The purpose of rule option \CodeIn{msg}, for
instance, is to print on the output an informative message indicating
what kind of intrusion has been
observed. Figure~\ref{fig:distribution-rules-protocol} shows the
distributions of these options per protocol and their relevance for
triggering rule actions. We obtained this information
by\Fix{...explicar como...} \Fix{eh um pouco suspeito ter tantas
  regras irrelevantes...}

%% This plot indicates that the problem search
%% space is defined based on the protocol of the malicious traffic. For
%% instance, if \tname{} detects by analyzing the format of network
%% packets that the attack involves \CodeIn{http} traffic, then 34
%% different rule options will be considered in the creation of the rule.

Table~\ref{table:rules}
shows a sample of the relevant options of Suricata. Column ``\#''
shows the option number, column ``Name'' shows the name of the
option, column ``\#~Params'' shows the number of additional parameters
associated with a given option, and column ``Description'' presents a
short description of the option purpose. \Fix{...explain content...}

@@

@@

The SYN Flood attack is a denial-of-service attack that exploits a vulnerability in the TCP/IP handshake
to establish a TCP connection~\cite{cloudfare-synflood}. The handshake
works as follows in normal circumstances. First, a client sends a SYN
packet to the server, requesting a connection. Second, the server
responds with a SYN-ACK packet to the client. Third, the client
responds with an ACK message and the connection is established. Aware
of the protocol, an attacker sends multiple SYN packets to different
ports of a server, often using fake IP addresses. Then, after the
server responds with a SYN-ACK packet, the client keeps sending other
SYN packets to avoid the connection to time out. Without proper
protection, the server accepts these malicious requests and eventually
legitimate requests cannot be satisfied due to resource exhaustion.

\begin{figure}[h!]
  \lstinputlisting[language=C,numbers=none,keywords={flags,threshold}]{synflood.suricata}
  \caption{Suricata rule for SYN Flood Attacks.}
  \label{fig:synflood-example}
\end{figure}

Figure~\ref{fig:synflood-example} shows a Suricata rule that captures
this attack. This rule was obtained from the \Fix{...what source...}
The relevant options in the rule appear in bold---the option
\CodeIn{flags: S,12}, identifies a SYN packet in a TCP packet and the
option \CodeIn{threshold: type both, track by\_dst, count 5000,
  seconds 5} indicates that a high volume of such packets should be
requested in a short period of time.

@@

@@

\begin{figure*}[t]
\centering
\includegraphics[trim=0 300 350 0,clip,width=0.95\textwidth]{figs/nids-workflow}
\caption{\tname\ workflow.}
\label{fig:overview}
\end{figure*}

\vspace{1ex}
\noindent\textbf{Overview.}~ The input of the technique is the
positive traffic whereas the output is a list of minimized rules
capturing only the positive, malicious,
traffic. Figure~\ref{fig:overview} shows the workflow of \tname{} as a
pipeline of two components, which appear numbered in the figure.
\tname{} synthesizes rules in two steps. First, it produces a
potentially overspecified rule that captures the malicious traffic
provided on input. In this step, \tname\ finds options by
\emph{reverse engineering} the positive traffic. It extracts tokens
from the malicious message and uses them to define the initial
rule. We say that this rule overfits the problem as it satisfies the
success criterion of only capturing the positive traffic, however it
is unlikely that it will be able to generalize to other attacks (say,
attacks of the same kind, but with different data). The reason is that
many options added to the rule at the first step are not important for
discriminating the attack. In the second step, \tname{} uses negative
examples to \emph{minimize} the rule, discarding unnecessary
options. Note that any subset of the original set of options captures
the positive traffic. The minimization step uses a population-based
gradient descent-like search to generate minimal solutions.

%% The goal of the first
%% component is to produce a rule that captures the negative traffic.
%% First, it initializes the rule with random values. Only options
%% associated with the used protocol are included in the rule
%% representation. Then, \tname\ optimizes the rule options until
%% \suri\ is able to capture the attack associated with the negative
%% traffic.  \tname\ unsuccessfully terminates at this point if it cannot
%% capture the attack. The second component takes as input the rule
%% produced by the first component and tries to minimize that rule. Note
%% that any subset of options would capture the negative traffic---as all
%% options in the rule need to be satisfied for the negative traffic to
%% be captured---but it can also capture positive traffic. This component
%% systematically discards options from the rule encoding until no
%% positive traffic is captured.

\subsection{Step 1: Reverse Engineering}
\pgfplotsset{width=6cm,compat=1.8}
\pgfplotsset{every tick label/.append style={font=\tiny}}

\Fix{--------------------------- estou aqui}

The goal of the first step is to extract options from the positive
traffic. To infer options, \tname{} parses the protocol message, looking for
fields in the message associated with options. For example, fields
\Fix{...and...}  can be used to determine the option
\Fix{...}. Unfortunately, inferring options at the field granularity
level is insufficient to capture \CodeIn{content} options (see
Section~\ref{sec:example-suricata-rules}), which refer to parts of the
payload of the message and cannot be directly extracted directly from
the fields. Handling this case is critically important as most rules
contain \CodeIn{content} options. Figure~\ref{fig:distribution-contents} shows
the histogram of number of these options per rule (up to 10) for the
Suricata public rule database \Fix{cite}. Rules without \CodeIn{content}
options is an exception and half of the rules contain at least two
\CodeIn{content} options.
\begin{wrapfigure}[14]{r}{0.56\textwidth}
  \centering
  \vspace{-3ex}
  \scalebox{1.2}{  
    \begin{tikzpicture}
      \begin{axis}[
          bar width=6pt,
          scaled ticks=false,
          tick label style={/pgf/number format/fixed},
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          symbolic x coords={0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10},
          xtick=data,
          nodes near coords,
          every node near coord/.append style={font=\tiny},
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(0,25) (1,10222) (2,4956) (3,5589)
          (4,1651) (5,1457) (6, 729) (7, 471) (8, 313) (9, 137) (10, 126)};
      \end{axis}
    \end{tikzpicture}
  }
  \caption{\label{fig:distribution-contents}Histogram of number of
    \CodeIn{content} options per rule (up to 10).}
\end{wrapfigure}
To find the values associated with these options, \tname{} splits the
payload of the malicious message in tokens using natural language
delimiters, such as $\backslash$t, $\backslash$n, $\backslash$r, and
spaces. The rationale for this decision is that we observed, by
inspecting existing rules, that \CodeIn{content} options typically refer to
sequence of characters in the payload of messages separated by these
delimiters. \Fix{describe extraction of options not in the
  message. give examples! search only makes sense if this exists!}

@@
